---
title: "Data Munging for Data Visualization"
author: "Aaron R. Williams"
output:
  html_document:
    code_folding: show
    toc: TRUE
    toc_float: TRUE
    css: !expr here::here("www", "web_report.css")
editor_options:
  chunk_output_type: console
---

```{css echo=FALSE}
h1 {
    font-size: 34px;
    color: #337AB7;
}
p {
    margin: 20px 0 20px;
}
```

```{r message = FALSE, warning = FALSE}
library(tidyverse)

```

## Import Packages and Functions

* `filter()` - drop cases based on logical conditions
* `mutate()` - add new variables are transform existing variables
* `pivot_wider()` - make data wider
* `pivot_longer()` - make data longer
* `separate()` - split a character variable into multiple columns based on a delimiter

## Tidy Data

This data comes from `library(tidyr)` and these examples largely come from chapter 12 in [R for Data Science](https://r4ds.had.co.nz/tidy-data.html) by Hadley Wickham and Garrett Grolemund.

> table1, table2, table3, table4a, table4b, and table5 all display the number of TB cases documented by the World Health Organization in Afghanistan, Brazil, and China between 1999 and 2000. 

The following table, `table1` is tidy because

1. Each variable has its own column
2. Each observation has its own row
3. Each value has its own cell

```{r}
table1

```

## Untidy table2 {.tabset}

### Untidy

**Why aren't the data tidy?**

```{r}
table2

```

### Cleaned

**Each observation was spread across two rows!**

```{r}
table2 %>%
  pivot_wider(names_from = type, values_from = count)

```

## Untidy table3 {.tabset}

### Untidy

**Why aren't the data tidy?**

```{r}
table3

```

### Cleaned

**The `rate` column had two variables!**

```{r}
table3 %>%
  separate(rate, into = c("cases", "population")) %>%
  mutate(
    cases = as.numeric(cases),
    population = as.numeric(population)
  )

```

## Untidy table4 {.tabset}

### Untidy

**Why aren't the data tidy?**

```{r}
table4a
table4b

```

### Cleaned

**Each table had variables in the column names and the data were spread across two tables**

```{r}
# fix table4a
table4a_fixed <- table4a %>%
  pivot_longer(
    cols = c(`1999`, `2000`), 
    names_to = "year", 
    values_to = "cases"
  )

# fix table4b
table4b_fixed <- table4b %>%
  pivot_longer(
    cols = c(`1999`, `2000`), 
    names_to = "year", 
    values_to = "population"
  )

# join the two tables into one tidy table
left_join(
  table4a_fixed, 
  table4b_fixed, 
  by = c("country", "year")
)

```

## Titanic Data

Consider the Titanic data set from the 3D plots example. 

```{r eval = FALSE}
tribble(
  ~Class, ~Sex, ~n,
  "1st class", "female passengers", 144,
  "1st class", "male passengers", 179,
  "2nd class", "female passengers", 106,
  "2nd class", "male passengers", 171, 
  "3rd class", "female passengers", 216,
  "3rd class", "male passengers", 493
)


```

### `r kableExtra::text_spec("Exercise 1", color = "#1696d2")`

Tidy the following data:

```{r eval = FALSE}
tribble(
  ~Class, ~female_passengers, ~male_passengers,
  "1st class", 144, 179,
  "2nd class", 106, 171,
  "3rd class", 216, 493
)

```

### `r kableExtra::text_spec("Exercise 2", color = "#1696d2")`

Tidy the following data:

```{r eval = FALSE}
tribble(
  ~Class, ~`male passengers/female passengers`,
  "1st class", "179|144",
  "2nd class", "171|106",
  "3rd class", "493|216",
)

```

### `r kableExtra::text_spec("Exercise 3", color = "#1696d2")`

Suppose we want to create a column chart with only female passengers. Furthermore, we want `n` on the x-axis and we want `Class` on the y-axis with the levels going from 1st to 3rd from top to bottom.

<font color="#55b748">Step 1:</font> Use `filter()` to drop male passengers. 

<font color="#55b748">Step 2:</font> Create a column chart with `n` on the x-axis and `Class` on the y-axis. Pipe (`%>%`) the data into `ggplot()`. 

<font color="#55b748">Step 3:</font> After the `filter()`, add the following:

```
  mutate(Class = factor(Class, levels = c("3rd class", "2nd class", "1st class"))) %>%

```

## Additional tools

* `library(stringr)` has powerful tools for dealing with text strings.
* `library(lubridate)` has powerful tools for dealing with time and dates.
